---
title: "Life Table Models"
author: "Jared Bennett"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
#output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1) Life Table
The purpose of this file is to test the fits of different distributions to 
human life tables. Distributions of interest are the Gompertz and Weibull. I will 
also fit multiple splines, as we really only care about accurate fits.  


```{r, eval=TRUE, echo=TRUE, results='hold'}
invisible(rm(list=ls()));invisible(gc());
###############################################################################
###############################################################################
# Import data and view
USPop <- read.csv(file = "~/Desktop/LifeTables/1-TanzaniaPop.csv", header = TRUE, stringsAsFactors = FALSE)

par(las = 1, font.lab = 2, font.axis = 2, font.main = 2, cex.main = 1.75,
    mar=c(4.1,4.1,4.1,2.1), mgp = c(2.5,.75,0))

plot(x = USPop[ ,1], y = USPop[ ,2], type = "p", pch = 16, ylim = c(0,1),
     main = "Us Population Death Proportion", ylab = "Fraction of Death", lwd=2,
     xlab = "Age (Years)")
box(lwd = 2)
grid()


par(las = 1, font.lab = 2, font.axis = 2, font.main = 2, cex.main = 1.75,
    mar=c(4.1,4.1,4.1,2.1), mgp = c(2.5,.75,0))

plot(x = USPop[ ,1], y = log(USPop[ ,2]), type = "p", pch = 16,
     main = "Us Population Death Proportion", ylab = "log(Fraction of Death)", lwd=2,
     xlab = "Age (Years)")
box(lwd = 2)
grid()
###############################################################################
###############################################################################
# plot
# testLabels <- c(13:52, 1:32)
# 
# par(las = 1, font.lab = 2, font.axis = 2, font.main = 2, cex.main = 1.75,
#     mar=c(4.1,4.1,4.1,2.1), mgp = c(2.5,.75,0))
# 
# plot(x = Chikungunya[ ,1], y = Chikungunya[ ,2]/N, type = "p", pch = 16, ylim = c(0,0.15),
#      main = "Chikungunya Incidence Data", ylab = "Fractional Incidence", lwd=2,
#      xlab = "Week", xaxt = 'n')
# lines(x = Chikungunya[ ,1], y = initMod, type = "l", lwd=2, col = "Magenta")
# axis(side = 1, at = Chikungunya[,1][c(T,F)], labels = testLabels[c(T,F)])
# box(lwd = 2)
# grid()
# legend(x = "topright", legend = c("Incidence Data", "Model Estimate"), col = c("black", "magenta"),
#        bty = "n", lty = 1, lwd = 3, cex = 1)

```


#2) Models To Fit
I want to test the Gompertz distribution, the Weibull distribution, and splines 
of degrees 2 through 5.

```{r, eval=TRUE, echo=TRUE, results='hold'}
###############################################################################
###############################################################################
# distribution functions
dGompertz <- function(x, p){
  # p[1] is frailty, p[2] is scale
  exponent <- p[1]*(1-exp(p[2]*x))
  total <- p[2]*x + exponent
  return(p[2]*p[1]*exp(total))
}

dWeibull <- function(x,p){
  #p[1] is shape, p[2] is scale
  exponent1 <- exp(-(x/p[2])^p[1])
  exponent2 <- (x/p[2])^(p[1]-1)
  return(p[1]/p[2]*exponent2*exponent1)
}

# splines
spline_2 <- function(x, p) return(p[1]*x + p[2]*x^2)
spline_3 <- function(x, p) return(p[1]*x + p[2]*x^2 + p[3]*x^3)
spline_4 <- function(x, p) return(p[1]*x + p[2]*x^2 + p[3]*x^3 + p[4]*x^4)
spline_5 <- function(x, p) return(p[1]*x + p[2]*x^2 + p[3]*x^3 + p[4]*x^4 + p[5]*x^5)

# splines with shifted exponential
spline1Exp <- function(x, p) return(p[1] + p[2]*x + p[3]*exp(x-100.5))
spline2Exp <- function(x, p) return(p[1] + p[2]*x + p[3]*x^2 + p[4]*exp(x-x[length(x)]))
spline3Exp <- function(x, p) return(p[1] + p[2]*x + p[3]*x^2 + p[4]*x^3 + p[5]*exp(x-x[length(x)]))
spline4Exp <- function(x, p) return(p[1] + p[2]*x + p[3]*x^2 + p[4]*x^3 + p[5]*x^4 + p[6]*exp(x-x[length(x)]))
spline5Exp <- function(x, p) return(p[1] + p[2]*x + p[3]*x^2 + p[4]*x^3 + p[5]*x^4 + p[6]*x^5 + p[7]*exp(x-x[length(x)]))

# splines with shifted exponential and default y=0 at x=0
spline1Exp0 <- function(x, p) return(-exp(-x[length(x)]) + p[1]*x + p[2]*exp(x-x[length(x)]))
spline2Exp0 <- function(x, p) return(-exp(-x[length(x)]) + p[1]*x + p[2]*x^2 + p[3]*exp(x-x[length(x)]))
spline3Exp0 <- function(x, p) return(-exp(-x[length(x)]) + p[1]*x + p[2]*x^2 + p[3]*x^3 + p[4]*exp(x-x[length(x)]))
spline4Exp0 <- function(x, p) return(-exp(-x[length(x)]) + p[1]*x + p[2]*x^2 + p[3]*x^3 + p[4]*x^4 + p[5]*exp(x-x[length(x)]))
spline5Exp0 <- function(x, p) return(-exp(-x[length(x)]) + p[1]*x + p[2]*x^2 + p[3]*x^3 + p[4]*x^4 + p[5]*x^5 + p[6]*exp(x-x[length(x)]))

# Just in case my assumptions suck, and because it's easy to copy/paste
spline1ExpD <- function(x, p) return(-p[1] + p[2]*x + p[3]*exp(x-p[4]))
spline2ExpD <- function(x, p) return(-p[1] + p[2]*x + p[2]*x^2 + p[4]*exp(x-p[5]))
spline3ExpD <- function(x, p) return(-p[1] + p[2]*x + p[3]*x^2 + p[4]*x^3 + p[5]*exp(x-p[6]))
spline4ExpD <- function(x, p) return(-p[1] + p[2]*x + p[3]*x^2 + p[4]*x^3 + p[5]*x^4 + p[6]*exp(x-p[7]))
spline5ExpD <- function(x, p) return(-p[1] + p[2]*x + p[3]*x^2 + p[4]*x^3 + p[5]*x^4 + p[6]*x^5 + p[7]*exp(x-p[8]))

###############################################################################
###############################################################################
# Cost function
costFunc <- function(params, testFunc, data){

  estimates <- testFunc(x = data[ ,1], p = params)
  
  #least squares
  return( sum((data[ ,2] - estimates)^2) )
}
###############################################################################
###############################################################################
# Model scoring
AIC_LS <- function(n, k, RSS){
  #n: number of observations. 101 here
  #k: number of parameters. here, length(p)
  #RSS: Residual sum of squares. here, output of costFunc and optim$value
  return(2*k + n*log(x = RSS))
}

BIC_LS <- function(n, k, RSS){
  #n: number of observations. 101 here
  #k: number of parameters. here, length(p)
  #RSS: Residual sum of squares. here, output of costFunc and optim$value
  return(log(x = n)*k + n*log(x = RSS))
}
```


#3) Setup Data and Fit

```{r, eval=TRUE, echo=TRUE, results='hold'}
###############################################################################
###############################################################################
# Setup functions list and parameters lists
funcList <- list(dGompertz, dWeibull, spline_2, spline_3, spline_4, spline_5,
                 spline1Exp, spline2Exp, spline3Exp, spline4Exp, spline5Exp,
                 spline1Exp0, spline2Exp0, spline3Exp0, spline4Exp0, spline5Exp0,
                 spline1ExpD, spline2ExpD, spline3ExpD, spline4ExpD, spline5ExpD)

paramList1 <- c(lapply(X = c(2,2,2:5), FUN = rep, x = 0.1),
                lapply(X = 3:7, FUN = rep, x = 0.1),
                lapply(X = 2:6, FUN = rep, x = 0.1),
                lapply(X = lapply(X = 3:7, FUN = rep, x = 0.1), FUN = 'c', 100))

paramList2 <- c(lapply(X = c(2,2,2:5), FUN = rep, x = 0.01),
                lapply(X = 3:7, FUN = rep, x = 0.01),
                lapply(X = 2:6, FUN = rep, x = 0.01),
                lapply(X = lapply(X = 3:7, FUN = rep, x = 0.01), FUN = 'c', 100))

paramList3 <- c(lapply(X = c(2,2,2:5), FUN = rep, x = 0.001),
                lapply(X = 3:7, FUN = rep, x = 0.001),
                lapply(X = 2:6, FUN = rep, x = 0.001),
                lapply(X = lapply(X = 3:7, FUN = rep, x = 0.01), FUN = 'c', 100))

paramList4 <- c(lapply(X = c(2,2,2:5), FUN = rep, x = 0.0001),
                lapply(X = 3:7, FUN = rep, x = 0.0001),
                lapply(X = 2:6, FUN = rep, x = 0.0001),
                lapply(X = lapply(X = 3:7, FUN = rep, x = 0.01), FUN = 'c', 100))

nDeps <- lapply(X = lapply(X = paramList4, FUN = length), FUN = rep, x = 1e-5)


###############################################################################
###############################################################################
# fit things
fit1 <- fit2 <- fit3 <- fit4 <- vector(mode = "list", length = length(funcList))

for(iter in 1:length(funcList)){
  
  fit1[[iter]] <- optim(par = paramList1[[iter]], fn = costFunc, testFunc = funcList[[iter]],
                        data = USPop, method = "BFGS", control = list(maxit = 1000, ndeps = nDeps[[iter]]))
  
  fit2[[iter]] <- optim(par = paramList2[[iter]], fn = costFunc, testFunc = funcList[[iter]],
                        data = USPop, method = "BFGS", control = list(maxit = 1000, ndeps = nDeps[[iter]]))
  
  fit3[[iter]] <- optim(par = paramList3[[iter]], fn = costFunc, testFunc = funcList[[iter]],
                        data = USPop, method = "BFGS", control = list(maxit = 1000, ndeps = nDeps[[iter]]))

  fit4[[iter]] <- optim(par = paramList4[[iter]], fn = costFunc, testFunc = funcList[[iter]],
                        data = USPop, method = "BFGS", control = list(maxit = 1000, ndeps = nDeps[[iter]]))
}

###############################################################################
###############################################################################
# get AIC and BIC values

AIC1 <- AIC2 <- AIC3 <- AIC4 <- numeric(length = length(funcList))
BIC1 <- BIC2 <- BIC3 <- BIC4 <- numeric(length = length(funcList))

numObs <- dim(USPop)[1]

for(iter in 1:length(funcList)){
  # get things used several times
  numK <- length(paramList1[[iter]])
  
  # AIC calcs
  AIC1[iter] <- AIC_LS(n = numObs, k = numK, RSS = fit1[[iter]]$value)
  AIC2[iter] <- AIC_LS(n = numObs, k = numK, RSS = fit2[[iter]]$value)
  AIC3[iter] <- AIC_LS(n = numObs, k = numK, RSS = fit3[[iter]]$value)
  AIC4[iter] <- AIC_LS(n = numObs, k = numK, RSS = fit4[[iter]]$value)
  
  # BIC calcs
  BIC1[iter] <- BIC_LS(n = numObs, k = numK, RSS = fit1[[iter]]$value)
  BIC2[iter] <- BIC_LS(n = numObs, k = numK, RSS = fit2[[iter]]$value)
  BIC3[iter] <- BIC_LS(n = numObs, k = numK, RSS = fit3[[iter]]$value)
  BIC4[iter] <- BIC_LS(n = numObs, k = numK, RSS = fit4[[iter]]$value)
  
}

```






#4) Plot Results
```{r, eval=TRUE, echo=TRUE, results='hold'}
###############################################################################
###############################################################################
# plots of data!!!

# labels used for everything
xDim <- length(funcList)
xPoints1 <- jitter(x = 1:xDim, amount = .08)
xPoints2 <- jitter(x = 1:xDim, amount = .08)
xPoints3 <- jitter(x = 1:xDim, amount = .08)
xPoints4 <- jitter(x = 1:xDim, amount = .08)

testLabels <- c("dGompertz", "dWeibull", "spline_2", "spline_3", "spline_4", "spline_5",
                 "spline1E", "spline2E", "spline3E", "spline4E", "spline5E",
                 "spline1E0", "spline2E0", "spline3E0", "spline4E0", "spline5E0",
                 "spline1ED", "spline2ED", "spline3ED", "spline4ED", "spline5ED")


#######################################
# Plot Residuals
fit1Vals <- log(vapply(X = fit1, FUN = '[[', 2, FUN.VALUE = numeric(length = 1L)))
fit2Vals <- log(vapply(X = fit2, FUN = '[[', 2, FUN.VALUE = numeric(length = 1L)))
fit3Vals <- log(vapply(X = fit3, FUN = '[[', 2, FUN.VALUE = numeric(length = 1L)))
fit4Vals <- log(vapply(X = fit4, FUN = '[[', 2, FUN.VALUE = numeric(length = 1L)))

ybounds <- 1.1*c(min(fit1Vals,fit2Vals,fit3Vals,fit4Vals),
                 max(fit1Vals,fit2Vals,fit3Vals,fit4Vals))

# set output
png(filename = "./10_Residuals_Full.png", width=1920, height=1080, units="px")

# plot
par(las = 1, font.lab = 2, font.axis = 2, font.main = 2, cex.main = 4,
    cex.lab = 3, cex.axis = 2, mar=c(10,8,5,2), mgp = c(5,1,0))

plot(x = xPoints1, y = fit1Vals, type = "p", pch = 16, ylim = ybounds,
     main = "Residuals For All Model Fits", ylab = "ln(Residuals)", lwd=2,
     xlab = "", xaxt = 'n', cex = 2)
points(x = xPoints2, y = fit2Vals, type = "p", pch = 16, col="red", cex = 2)
points(x = xPoints3, y = fit3Vals, type = "p", pch = 16, col="green", cex = 2)
points(x = xPoints4, y = fit4Vals, type = "p", pch = 16, col="purple", cex = 2)

axis(side = 1, at = 1:xDim, labels = FALSE)
text(x = 1:xDim, y = par("usr")[3] - 0.25, srt = 60, adj = c(0.75,-0.5),
     labels = testLabels, xpd = TRUE, font = 2, cex = 2)

box(lwd = 2)
grid()

legend(x = "top", legend = c("Fit1", "Fit2", "Fit3", "Fit4"),
       col = c("black", "red", "green", "purple"),
       bty = "n", lty = 1, lwd=5, cex = 2, text.font = 2)

## close output to file
dev.off()


#######################################
# Plot AIC
ybounds <- 1.1*c(min(unlist(AIC1),unlist(AIC2),unlist(AIC3),unlist(AIC4)),
                 max(unlist(AIC1),unlist(AIC2),unlist(AIC3),unlist(AIC4)))

fileNames <- c("./20_AIC_Full.png", "./21_AIC_Zoom.png")
ybounds <- list(ybounds, c(ybounds[1], -ybounds[1]))

for(i in 1:1){
  # set output
  png(filename = fileNames[i], width=1920, height=1080, units="px")
  
  # plot
  par(las = 1, font.lab = 2, font.axis = 2, font.main = 2, cex.main = 4,
      cex.lab = 3, cex.axis = 2, mar=c(10,8,5,2), mgp = c(5,1,0))

  plot(x = xPoints1, y = unlist(AIC1), type = "p", pch = 16, ylim = ybounds[[i]],
       main = "AIC For All Model Fits", ylab = "AIC", lwd=2,
       xlab = "", xaxt = 'n', cex = 2)
  points(x = xPoints2, y = unlist(AIC2), type = "p", pch = 16, col="red", cex = 2)
  points(x = xPoints3, y = unlist(AIC3), type = "p", pch = 16, col="green", cex = 2)
  points(x = xPoints4, y = unlist(AIC4), type = "p", pch = 16, col="purple", cex = 2)
  
  axis(side = 1, at = 1:xDim, labels = FALSE)
  text(x = 1:xDim, y = par("usr")[3] - 0.25, srt = 60, adj = c(1.15,1.2),
       labels = testLabels, xpd = TRUE, font = 2, cex = 2)
  
  box(lwd = 2)
  grid()
  
  legend(x = "top", legend = c("AIC1", "AIC2", "AIC3", "AIC4"),
         col = c("black", "red", "green", "purple"),
         bty = "n", lty = 1, lwd=5, cex = 2, text.font = 2)
  
  ## close output to file
  dev.off()
}

#######################################
# Plot BIC
ybounds <- 1.1*c(min(unlist(BIC1),unlist(BIC2),unlist(BIC3),unlist(BIC4)),
                 max(unlist(BIC1),unlist(BIC2),unlist(BIC3),unlist(BIC4)))

fileNames <- c("./30_BIC_Full.png", "./31_BIC_Zoom.png")
ybounds <- list(ybounds, c(ybounds[1], -ybounds[1]))

for(i in 1:1){
  # set output
  png(filename = fileNames[i], width=1920, height=1080, units="px")
  
  # plot
  par(las = 1, font.lab = 2, font.axis = 2, font.main = 2, cex.main = 4,
      cex.lab = 3, cex.axis = 2, mar=c(10,8,5,2), mgp = c(5,1,0))

  plot(x = xPoints1, y = unlist(BIC1), type = "p", pch = 16, ylim = ybounds[[i]],
       main = "BIC For All Model Fits", ylab = "BIC", lwd=2,
       xlab = "", xaxt = 'n', cex = 2)
  points(x = xPoints2, y = unlist(BIC2), type = "p", pch = 16, col="red", cex = 2)
  points(x = xPoints3, y = unlist(BIC3), type = "p", pch = 16, col="green", cex = 2)
  points(x = xPoints4, y = unlist(BIC4), type = "p", pch = 16, col="purple", cex = 2)
  
  axis(side = 1, at = 1:xDim, labels = FALSE)
  text(x = 1:xDim, y = par("usr")[3] - 0.25, srt = 60, adj = c(1.15,1.2),
       labels = testLabels, xpd = TRUE, font = 2, cex = 2)
  
  box(lwd = 2)
  grid()
  
  legend(x = "top", legend = c("BIC1", "BIC2", "BIC3", "BIC4"),
         col = c("black", "red", "green", "purple"),
         bty = "n", lty = 1, lwd=5, cex = 2, text.font = 2)
  
  ## close output to file
  dev.off()
}

#######################################
# Plot Ranks

# get rank of each run
orderAIC <- lapply(X = list(AIC1,AIC2,AIC3,AIC4), FUN = rank)
orderBIC <- lapply(X = list(BIC1,BIC2,BIC3,BIC4), FUN = rank)
orderFit <- lapply(X = list(fit1Vals, fit2Vals, fit3Vals, fit4Vals), FUN = rank)

#get mean rank
orderAICMean <- Reduce(f = "+", x = orderAIC)/4
orderBICMean <- Reduce(f = "+", x = orderBIC)/4
orderFitMean <- Reduce(f = "+", x = orderFit)/4

# set output
png(filename = "./40_RankOrder.png", width=1920, height=1080, units="px")

# plot1
par(mfrow = c(1,2), las = 1, font.lab = 2, font.axis = 2, font.main = 2,
    cex.main = 4, cex.lab = 3, cex.axis = 2, mar=c(6,7,5,2), mgp = c(4,1,0))

plot(x = orderFitMean, y = orderAICMean, type = "p", pch = 16, ylim = c(0,22),
     xlim = c(0,22),main = "Model Fit vs AIC", ylab = "AIC Mean Rank",
     lwd=2, xlab = "Residual Mean Rank", cex = 2)

box(lwd = 2)
grid()

text(x = orderFitMean, y = orderAICMean,
     labels = testLabels[order(orderAICMean)][rank(orderAICMean)],
     pos = 4, offset = 0.5)

# plot2
plot(x = orderFitMean, y = orderBICMean, type = "p", pch = 16, ylim = c(0,22),
     xlim = c(0,22), main = "Model Fit vs BIC", ylab = "BIC Mean Rank",
     lwd=2, xlab = "Residual Mean Rank", cex = 2)

box(lwd = 2)
grid()

text(x = orderFitMean, y = orderBICMean,
     labels = testLabels[order(orderBICMean)][rank(orderBICMean)],
     pos = 4, offset = 0.5)

## close output to file
dev.off()

#######################################
# Plot fit

# set output
png(filename = "./50_FitWithData.png", width=1920, height=1080, units="px")

# plot1
par(las = 1, font.lab = 2, font.axis = 2, font.main = 2,
    cex.main = 4, cex.lab = 3, cex.axis = 2, mar=c(6,7,5,2), mgp = c(4,1,0))

plot(x = USPop[ ,1], y = USPop[ ,2], type = "p", pch = 16, ylim = c(0,1),
     main = "US Population Death Proportion", ylab = "Fraction of Death", lwd=2,
     xlab = "Age (Years)")

lines(x = USPop[ ,1], y = spline5Exp(x = USPop[ ,1], p = fit4[[11]]$par),
      type = "l", col = "magenta", lwd = 2)

box(lwd = 2)
grid()

legend(x = "top", legend = c("Data", "spline_5_Exp"),
       col = c("black", "magenta"),
       bty = "n", lty = 1, lwd=5, cex = 2, text.font = 2)

## close output to file
dev.off()
```



#4) Final Function to Add to Package
```{r, eval=TRUE, echo=TRUE, results='hold'}
#' Fit Life Table Data
#'
#' This function reads in a life table of 2 columns (see examples in the data), 
#' where the first column is the mean age of the published data (ie, if published 
#' data is 1-2 years, the input file has 1.5) and the second column is q_x, the probability 
#' of dying during that time period. The output is the complete fit output, from which 
#' parameters can be taken. Least squares is used as the cost function, and base::optim
#' is used to perform fits. A 5th-order spline with an exponential is used to fit against.
#' 
#' @param lifeTable Path to .csv life table where the first two columns are age and qx
#' @param initParams Initial estimates for coefficients. Defaults to 0.001 for all
#' @param method Method to feed to optim. Default is BFGS, which performed best in all tests
#' @param fitArgs List of args to feed to optim. Defaults include the max iterations and step size
#'
#' @examples
#' results = FitLifeTable(lifeTable = "~/PATHTOFILE.csv")
#' coefficients = results$par
#'
#' @return Named list of the same structure as optim. Parameters can be obtained 
#' by return$par
#'
#' @export
FitLifeTable <- function(lifeTable,
                         initParams = rep(x = 0.001, 7),
                         method = "BFGS",
                         fitArgs = list(maxit = 500, ndeps = rep(x = 1e-5, 7))
                         ){
  
  # import data to fit against
  lifeData <- read.csv(file = lifeTable, header = TRUE, stringsAsFactors = FALSE)
  
  # this is the function we use for describing death probs at each age
  spline5Exp <- function(x, p) return(p[1] + p[2]*x + p[3]*x^2 + p[4]*x^3 + p[5]*x^4 + p[6]*x^5 + p[7]*exp(x-x[length(x)]))
  
  # Cost function - Least squares
  costFunc <- function(params, testFunc, data){
    #fit function against data with current params
    estimates <- testFunc(x = data[ ,1], p = params)
    #least squares
    return( sum((data[ ,2] - estimates)^2) )
  }

  # perform optimization
  results <- optim(par = initParams, fn = costFunc, testFunc = spline5Exp,
                   data = lifeData, method = method, control = fitArgs)

  return(results)
}


FitLifeTable(lifeTable = "../LifeTables/1-TanzaniaPop.csv")





```